---
title: Visualizing Mahomes' TD Passes
author: Jason Houle
date: '2019-02-12'
slug: visualizing-td-passes
summary: A simple visualization of a remarkable feat.
categories:
  - football
tags:
  - football
image:
  caption: 'Image credit: [**@Chiefs**](https://twitter.com/Chiefs/status/1081271078423531520)'
  focal_point: ''
output:
  blogdown::html_page:
    highlight: default
---

As a lifelong Kansas City Chiefs fan, I went from pleasantly surprised to awestruck to completely incredulous as I watched Patrick Mahomes' breakout 2018 NFL season. As his MVP-level performance continued game after game, I began to wonder: How does this feat compare with some of the all-time great season performances in TD passing? I decided to pull together some data quickly to compare.

```{r, results=FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(curl)
```

I scraped data from the [Pro Football Reference](https://www.pro-football-reference.com) single season touchdown leaders page, collecting the top seven performances for comparison.

```{r, cache=TRUE}
main_url <- "https://www.pro-football-reference.com/leaders/pass_td_single_season.htm"
url_root <- "https://www.pro-football-reference.com"

# Scrape records table
front_page <- read_html(curl(main_url))
a <- front_page %>%
  html_node(xpath = '//*[@id="leaders"]') %>% 
  html_table()
url <- front_page %>% 
  html_nodes(".left:nth-child(5) a") %>% 
  html_attr("href") %>% 
  paste0(url_root, .)

# Build tibble
qbs <- bind_cols(a, list(url = url)) %>% 
  head(7) %>%
  mutate(Player = trimws(str_extract(Player, "^[A-z ]*")),
         id = paste0(Tm, Year))

qbs
```

I wanted to break down the season game-by-game, and so I needed to scrape season pages. This proved a little trickier than I had hoped; while team season pages include a consolidated touchdown log, that field is apparently dynamically populated as the page loads, so it can't be retrieved with a `curl` command. Instead, I needed to access each boxscore link and scrape the number of touchdown passes in the game. (I figure there are alternatives using [RSelenium](https://cran.r-project.org/web/packages/RSelenium/) or a different data source like [`nflscrapR`](https://github.com/maksimhorowitz/nflscrapR), but I was too far down this path to switch!) For each season I built the cumulative TD pass totals by game as a data frame, and then within the `for` loop I appended these to create a list of dataframes, which were then added as a [list column](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html) to my main dataframe.

```{r}
# Scrape each QB season page
season <- vector("list", length(qbs[["id"]]))

for (i in seq_along(qbs[["url"]])) {
  num_games <- 16
  
  page_html <- read_html(curl(qbs[["url"]][i]))
  game_links <- page_html %>% 
    html_nodes(".center a") %>%
    html_attr("href") %>% 
    head(num_games)
  
  game <- 1:num_games
  tds <- vector("integer", num_games)
  
  for (g in seq_along(game_links)) {
    scoring <- read_html(curl(paste0(url_root, game_links[g]))) %>%
      html_node("#scoring") %>%
      html_table()
    tds[g] <- length(grep(paste0("yard pass from ", qbs$Player[i]), scoring$Detail))
  }
  
  season[[i]] <- data_frame(game = game, tds = cumsum(tds))
  
}

qbs <- qbs %>% 
  bind_cols(data_frame(season = season)) %>% 
  as_tibble()
```

Inspired by another [visualization of athletic prowess](https://www.swissinfo.ch/eng/business/age-of-success_federer-s-path-to-18-grand-slam-tennis-titles/42919320), I decided to create a step chart. With a quick unnest of the list column (and a quick call to `mutate` to add the year since Peyton Manning shows up in this list twice!), I whipped up a first attempt, and...

```{r, cache=TRUE}
qbs %>% 
  unnest(season) %>% 
  mutate(Player = as.factor(paste(Player, Year))) %>% 
  ggplot(aes(x = game, y = tds, color = Player)) +
  geom_step()
```

The output was... underwhelming. Most pressing is the line overlap: since each vertical line occurs at an integer, there is a lot of overlap of vertical lines (and many horizontal lines are also obscured, for example in week 14), making it difficult to track one player's performance. Additionally, it would definitely boost interpretability to order the players' appearance in the legend by their season-end totals (which should correspond to their position on the plot, at least by Week 16), and to include those totals in the legend. Finally, for both readability and authenticity, I wanted to use colors that correspond to the player's team rather than the `ggplot2` defaults.

I started by pulling NFL team colors from [this website](https://teamcolorcodes.com/nfl-team-color-codes/).

```{r}
palette <- tribble(
  ~Tm, ~color,
  "DEN", "#FB4F14",
  "NWE", "#B0B7BC",
  "KAN", "#E31837",
  "IND", "#002C5F",
  "MIA", "#008E97",
  "NOR", "#D3BC8D",
  "GNB", "#203731"
)
```

`scale_color_manual` needs a named vector to use these, with names corresponding to the value assigned to the `color` aesthetic. This means I'll have to mutate the player names *first* and create the named vector from an intermediate data frame. While I'm already in the mutate call, I also apply a 'jitter' to both the `game` and `tds` values so that each line segment is slightly offset from any others it might be overlapping. I then create the named vector and finally plot the jittered data, using a thicker than default line size and adding some labels, the manual color scale, and a minimal theme.

```{r, cache=TRUE}
qbs %>%
  unnest(season) %>%
  mutate(Player = factor(paste(Player, Year, paste0("(",TD,")"), sep = ", ")) %>% fct_inorder(),
         game = game + .05 * as.integer(Player) - .2,
         tds = tds + .1 * as.integer(Player) - .4) -> qbs2

qbs2 %>% 
  left_join(palette, by = "Tm") %>% 
  distinct(Player, color) -> color_df

color_lookup <- setNames(color_df$color, as.character(color_df$Player))

qbs2 %>% 
  ggplot(aes(x = game, y = tds, color = Player)) +
  geom_step(size = .8) +
  ggtitle("Season Passing Touchdown Count by Game") +
  xlab("Game number") +
  ylab("Passing Touchdown Count") +
  scale_color_manual(values = color_lookup) +
  theme_minimal()
```

There might be a few opportunities for improvement yet, but I think these come with tradeoffs: for instance, the amount of jitter could be increased to further reduce overlap, but this would come at the cost of readability. Some of the colors could be made more dissimilar (for instance, Manning '04 and Rodgers), but this would sacrifice some authenticity of using the official team colors.

All in all, this visualization packs in a lot of data and offers a side-by-side comparison of how Mahomes' MVP season stacks up with some of the best historical performances.
